---
title: "Relationship between a set of variables and miles per gallon (mpg)"
author: "Maurice"
date: "16 March 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5)
options(width=120)
```

## Executive Summary

This  looks at a dataset of a collection of cars in order to explore the relationship between a set of variables and miles per gallon, mpg, (outcome). There is interest in the following two questions:

“Is an automatic or manual transmission better for MPG”

"Quantify the MPG difference between automatic and manual transmissions"

We'll take the following steps:

- Process the data for use
- Explore the data especially focussing on the two paramaters of interest (Transmission and MPG)
- Model selection; where we try different models to help us answer our questions
- Model examination; to see if our best model holds up to our standards
- A Conclusion where we answer the questions based on the data

## Processing

First change 'am' to  factor (0 = automatic, 1 = manual)
And make cylinders a factor as well (since it is not continious)

```{r}
library(ggplot2)
library(GGally)
library(dplyr)
library(ggfortify)
data(mtcars)
mtcarsFactors <- mtcars
mtcarsFactors$am <- as.factor(mtcarsFactors$am)
levels(mtcarsFactors$am) <- c("automatic", "manual")
mtcarsFactors$cyl <- as.factor(mtcarsFactors$cyl)
mtcarsFactors$gear <- as.factor(mtcarsFactors$gear)
mtcarsFactors$vs <- as.factor(mtcarsFactors$vs)
levels(mtcarsFactors$vs) <- c("V", "S")
```

## Exploratory data analyses 

Look at the dimensions and head of the dataset to get an idea
```{r}
# Res 1
dim(mtcarsFactors)
# Res 2
head(mtcarsFactors)
```


Now we take a look at the realtionship between the two parameters we are intereseted in. 

```{r}
# Figure 1
library(ggplot2)
p <- ggplot(mtcarsFactors, aes(am, mpg))
p + geom_boxplot(aes(fill = am))
```

This clearly shows that the manual transmissions have higher mpg's. There could be a bias in the dataset that we are overlooking. 
Before creating a model let us look at which paramters to include besides 'am'. 
We look at all correlations of parameters and take only those higher than the 'am' correlation.

```{r}
# Res 3
cors <- cor(mtcars$mpg, mtcars)
orderedCors <- cors[,order(-abs(cors[1,]))]
orderedCors
# Res 4
amPos <- which(names(orderedCors)=="am")
subsetColumns <- names(orderedCors)[1:amPos]
subsetColumns
# Figure 2
mtcarsFactors[,subsetColumns] %>% 
   ggpairs(
     mapping = ggplot2::aes(color = am), 
     upper = list(continuous = wrap("cor", size = 3)),
     lower = list(continuous = wrap("smooth", alpha=0.4, size=1), combo = wrap("dot"))
   )
```


## Model selection 

Now that we have seen that mpg has many other (stronger) correlations other than 'am' we can guess that a model predicting the mpg solely on this parameter will not be the most accurate model. Let's check this out.

First we start with a basic linear model
```{r}
# Res 5
basicFit <- lm(mpg ~ am, mtcarsFactors)
summary(basicFit)
```

The p-values are actually quite low and the R-squared is problematic.
Now go to the other side of the spectrum by fitting all parameters of mtcars


```{r}
# Res 6
totalFit <- lm(mpg ~ ., mtcarsFactors)
summary(totalFit)
```

The R-squared has improved but the p-values hardly show any significance. Perhaps this is due to overfitting. We  have to meet somewhere in the middle.
Let's iterate using the step method

```{r}
# Res 7
bestFit <- step(totalFit,direction="both",trace=FALSE)
summary(bestFit)
```

## Model examination

The resulting best model `mpg ~ wt + qsec + am` is actually dependent on the transmission (am), but also weight (wt) and 1/4 mile time (qsec). All have significant p-values
The R-squared is pretty good to (0.85)

Now let's look (amongst others) at the Residuals vs Fitted 

```{r}
# Figure 3
autoplot(bestFit)
```

The 'Normal Q-Q' plot looks ok, but the 'Residuals vs Fitted' and 'Scale-Location' both show worrisome trends.

## Conclusion

The question “Is an automatic or manual transmission better for MPG” can be answered because all models (#Res5, #Res6 and #Res7) show that, holding all other paramters constant, manual transmission will increase your MPG.

The question “Quantify the MPG difference between automatic and manual transmissions” is quite tricky.

Based on the 'bestFit' (#Res7) model `mpg ~ wt + qsec + am` we could conclude that (with a p < 0.05 confidence) cars with manual transmission have 2.9358 (say 3) more miles per gallon than automatic transmissions.
The model seems clean with a p < 0.05 and R squared of 0.85

The residuals vs fitted chart however warns us that there is something missing in our model.
The real problem is that we only have 32 observations to train on (#Res1) and those observations hardly have overlap on the parameters 'wt' and 'qsec' (amongst others) if we look at the diagonal in the matrix chart (#Figure2)

Although the conclusion that mpg has better performance on manual transmissions appears to be alright, we cannot conclude that this model will fit with more observations.

